\section{The {\tt WHILE} Language\timeestimation{15h}}
\label{sec:WHILE}
While most modern programming languages today are turing complete and as such
are "equally powerful". For simplicity however not all languages are created
equal, as section \ref{sec:Turing Completeness} will note. Lets therefore
introduce a simple, but expressive enough language called {\tt WHILE}.

\begin{table*}[h]
	\begin{align*}
		Expression \ni E, F::= \qquad
										& \mathtt{nil} & | \\
										& \mathtt{cons}\; E\; F & | \\
										& \mathtt{hd}\; E     & | \\
										& \mathtt{tl}\; E     & | \\
										& \mathtt{[}I \mathtt{](} E \mathtt{)}      & | \\
										& I \\
		Statement \ni S, T::= \qquad
										& I \mathtt{:=} E & | \\
										& \mathtt{while}\; E\; \mathtt{\{}\; S\; \mathtt{\}} & | \\
										& S \; \text{\tt{<newline>}}\; T \\
		Program \ni P, Q::= \qquad
										& I_1\; \mathtt{read}\; I_2\; \mathtt{\{}\; S\; \mathtt{\}}\; \mathtt{write}\; I_3 & | \\
										& Q  \; \text{\tt{<newline>}}\; P
	\end{align*}
\end{table*}

Finally there are procedures. Procedures read an input and write an output and
execute some statements in between. (picture of input: data, black box, output:
data)

Procedures can be called by their names in brackets and the argument (a data
expression). Now however that this is just a convenience, so as to make writing
a moderately simple WHILE programs possible. One could just write everything
out, but that would be cumbersome and not very helpful in understanding
computability or complexity.

% subsection The Elements (end)
\paragraph{Alternatives to the {\tt WHILE} construct} % (fold)
\label{par:Alternatives to the WHILE construct}
\lineofthought{
	\begin{itemize}
		\item Unrestricted Recursion
		\item Unbound Search ($\mu$)
		\item Applying transition rules as long as possible (Markov)
  \end{itemize}
}
% paragraph Alternatives to the WHILE construct (end)
\subsubsection{Natural Numbers} % (fold)
\label{ssub:Natural Numbers}
\lineofthought{Bei \cite{jones} sind Zahlen unär codiert, zerstört das nicht 
alle möglichen Eigenschaften, wie {\tt add} in $\log n$ Zeit? Gucken, ob 
nicht Binär in Complexity vorkommt.}

% subsubsection Natural Numbers (end)
% subsection cons (end)
